Marker imputation efficiency for Genotyping-By-Sequencing data in rice (\emph{Oryza sativa}) and alfalfa (\emph{Medicago sativa}) \thanks{Nelson Nazzicari and Filippo Biscarini contributed equally to the work \\ Corresponding author: Nelson Nazzicari \email{nelson.nazzicari@crea.gov.it}}

### 2016-05-04 Nazzicari_Imputation.tex
aggregato

\newcolumntype{C}[1]{>{\centering\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}

@@

\makeatletter{}\begin{table}
\centering
\caption[Descriptive statistics]{
Descriptive statistics of alfalfa (\emph{Medicago sativa L.}) 
and rice (\emph{Oryza sativa L.}) genotyping. MAF is average minor allele frequency; 
p(AA), p(AB) and p(BB) are the proportions of AA, AB and BB genotypes.}
\label{tab:descriptive_statistics}
\begin{tabular}{lccccccc}
\hline\noalign{\smallskip}
\noalign{\smallskip}\hline\noalign{\smallskip}
\textbf{Dataset} & \textbf{No. samples} & \textbf{No. SNPs} & \textbf{Missing Rate} & \textbf{MAF} & \textbf{p(AA)} & \textbf{p(AB)} & \textbf{p(BB)}\\
\noalign{\smallskip}\Xhline{3\arrayrulewidth}\noalign{\smallskip}
Alfalfa-PV & 124 & 32\,706 & 0.666 & 0.1724 & 0.690 & 0.274 & 0.035\\
Alfalfa-Med & 154 & 40\,734 & 0.596 & 0.1702 & 0.691 & 0.276 & 0.032\\
Rice & 437 & 166\,418 & 0.534 & 0.140 & 0.860 & - & 0.140 \\
\noalign{\smallskip}\hline
\end{tabular}
\end{table}

@@

\makeatletter{}
\begin{table}
\centering
\caption[Markers counts and missing rates]{
Total number of markers and resulting average missing rate for the four different allowed missing rate thresholds in alfalfa (two datasets) and rice.}
\label{tab:missing_rates}
\begin{tabular}{C{3cm}rccc}
\hline\noalign{\smallskip}
\noalign{\smallskip}\hline\noalign{\smallskip}
\textbf{Allowed missing rate threshold} &  & \textbf{Alfalfa-PV} & \textbf{Alfalfa-Med} & \textbf{Rice}\\
\noalign{\smallskip}\Xhline{3\arrayrulewidth}\noalign{\smallskip}
   \multirow{2}{*}{70\%} & Missing rate &  0.325 & 0.29  & 0.364\\
                           & Markers      & 13\,190  & 19\,986 & 109\,372\\    
\noalign{\smallskip}\hline
   \multirow{2}{*}{40\%} & Missing rate &  0.161 & 0.142 & 0.198\\
                           & Markers      & 7\,790 & 12\,931 & 58\,553\\    
\noalign{\smallskip}\hline
   \multirow{2}{*}{20\%} & Missing rate & 0.076 & 0.074 & 0.099\\
                           & Markers      & 4\,828 & 8\,962 & 29\,872\\
\noalign{\smallskip}\hline                           
   \multirow{2}{*}{10\%} & Missing rate & 0.046 & 0.045 & 0.049\\
                           & Markers      & 3\,405 & 6\,364 & 15\,060 \\
\noalign{\smallskip}\hline
\end{tabular}
\end{table}

@@

Previous studies, though not detailing a per-class breakdown of the imputation accuracy, provide interesting comparisons. In sugar beet, Biscarini et al.~\cite{biscarini_genome-enabled_2014} used Beagle to impute SNP-chip markers with partial within-chromosome/scaffold alignment, finding global accuracies ranging from 84\% to 80.9\% with 1\% to 20\% missing genotypes. Huang et al.~\cite{huang_efficient_2014} tested several algorithms, including Beagle, on simulated ordered GBS rice data with missing rates in the 30-60\% range and finding Beagle gave imputation accuracy consistently higher (+15-20\%) than KNNI. 
Finally, Swarts et al.~\cite{swarts_novel_2014} used ordered GBS maize data to compare Beagle with the FSFHap and FILLIN algorithms they developed. They mostly obtained high imputation accuracies in the three genotype classes, providing further evidence of the added value of having a reference genome assembly.

@@

\makeatletter{}\section{Conclusions}
\label{sec:conclusions}
Imputation of missing genotypes can be an effective technique also for GBS data.
The most accurate imputation methods resulted in a total amount of wrongly imputed genotypes near to zero in rice and slightly over 10\% in alfalfa. 
The proportion of imputation errors, however, varied dramatically among genotype classes, approaching very high levels with the worst methods on the minor homozygous class.
The structure of the genome and the maturity of the reference assembly play a role in the imputation efficiency, as indicated by the greater efficiency obtained in rice compared with alfalfa. While Beagle was preferable for rice, general data imputation methods performed significantly better in the absence of a reference genome. In particular, in alfalfa RFI and KNNI showed the lowest imputation errors in all classes, with KNNI to be preferred for computational efficiency. The alignment of markers to closely related species can help, but comes at the price of discarding markers that do not align, and with general methods still performing better.\\
The imputation accuracy tended to be relatively robust over increasing missing rates. However, KNNI showed a somewhat lower accuracy on high missingness scenarios, probably as a consequence of the ``curse of dimensionality''. In terms of computation requirements, all methods proved to be tractable over all problem complexities with a standard bioinformatics-lab computation infrastructure, except RFI (whose computational requirements increased exponentially with problem size and took up to 40 days with the complete rice dataset and maximum missing rate thresholds).\\

@@

\makeatletter{}\begin{acknowledgements}
The rice data used in this research paper were produced within the framework of the Italian national project ``RISINNOVA'' (grant n. 2010-2369), financially supported by the AGER Foundation..
The creation of the alfalfa data sets was funded by the projects ‘Genomic selection in alfalfa (GENALFA)’ funded by the Italian Ministry of Foreign Affairs and International Cooperation in the framework of the Italy-USA scientific cooperation program, the Italian share of the FP7-ArimNet project ‘Resilient, water- and energy-efficient forage and feed crops for Mediterranean agricultural systems (REFORMA)’ funded by the Italian Ministry of Agricultural and Forestry Policies.
\end{acknowledgements}

>>>>

\newcolumntype{C}[1]{>{\centering\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
\newcommand{\specialcell}[2][c]{%
  \begin{tabular}[#1]{@{}c@{}}#2\end{tabular}}

@@

\begin{table}
\centering
\caption[Descriptive statistics, markers counts, and missing rates]{
Descriptive statistics of alfalfa (\emph{Medicago sativa L.}) 
and rice (\emph{Oryza sativa L.}) genotyping. MAF is average minor allele frequency; 
p(AA), p(AB) and p(BB) are the proportions of AA, AB and BB genotypes.
Total number of markers and resulting average missing rate for all markers
and for four allowed thresholds of missing rate per markers.}
\label{tab:missing_rates}
\begin{tabular}{C{3cm}rrrr}
\hline\noalign{\smallskip}
\specialcell{\textbf{Allowed missing}\\ \textbf{rate per marker}}  & & \textbf{Alfalfa-PV} & \textbf{Alfalfa-Med} & \textbf{Rice}\\
\noalign{\smallskip}\Xhline{3\arrayrulewidth}\noalign{\smallskip}
\multirow{7}{*}{100\%}  & \textbf{No. samples}  & 124     & 154       & 437\\
                        & \textbf{MAF}          & 0.1724  & 0.1702    & 0.140 \\
                        & \textbf{p(AA)}        & 0.690   & 0.691     & 0.860\\
                        & \textbf{p(AB)}        & 0.274   & 0.276     & -\\
                        & \textbf{p(BB)}        & 0.035   & 0.032     & 0.140\\
                        & \textbf{Missing rate} & 0.666   & 0.596     & 0.534\\
                        & \textbf{Markers}      & 32\,706 & 40\,734   & 166\,418\\    
\noalign{\smallskip}\hline\noalign{\smallskip}
   \multirow{2}{*}{70\%}   & \textbf{Missing rate} &  0.325 & 0.29  & 0.364\\
                           & \textbf{Markers}      & 13\,190  & 19\,986 & 109\,372\\    
\noalign{\smallskip}\hline\noalign{\smallskip}
   \multirow{2}{*}{40\%}   & \textbf{Missing rate} &  0.161 & 0.142 & 0.198\\
                           & \textbf{Markers}      & 7\,790 & 12\,931 & 58\,553\\    
\noalign{\smallskip}\hline\noalign{\smallskip}
   \multirow{2}{*}{20\%}   & \textbf{Missing rate} & 0.076 & 0.074 & 0.099\\
                           & \textbf{Markers}      & 4\,828 & 8\,962 & 29\,872\\
\noalign{\smallskip}\hline\noalign{\smallskip}
   \multirow{2}{*}{10\%}   & \textbf{Missing rate} & 0.046 & 0.045 & 0.049\\
                           & \textbf{Markers}      & 3\,405 & 6\,364 & 15\,060 \\
\noalign{\smallskip}\hline
\end{tabular}
\end{table}

@@

@@

Previous studies, though not detailing a per-class breakdown of the imputation accuracy, provide interesting comparisons. In sugar beet, Biscarini et al.~\cite{biscarini_genome-enabled_2014} used Beagle to impute SNP-chip markers with partial within-chromosome/scaffold alignment, finding global accuracies ranging from 84\% to 80.9\% with 1\% to 20\% missing genotypes. Huang et al.~\cite{huang_efficient_2014} tested several algorithms, including Beagle, on simulated ordered GBS rice data with missing rates in the 30-60\% range, and found that Beagle gave imputation accuracy consistently higher (+15-20\%) than KNNI. 
Finally, Swarts et al.~\cite{swarts_novel_2014} used ordered GBS maize data to compare Beagle with the FSFHap and FILLIN algorithms they developed. They mostly obtained high imputation accuracies in the three genotype classes, providing further evidence of the added value of having a reference genome assembly.

@@

\makeatletter{}\section{Conclusions}
\label{sec:conclusions}
Imputation of missing genotypes can be an effective technique also for GBS data.
The most accurate imputation methods resulted in a total amount of wrongly imputed genotypes near to zero in rice and slightly over 10\% in alfalfa. 
The proportion of imputation errors, however, varied dramatically among genotype classes, approaching very high levels with the worst methods on the minor homozygous class.
The structure of the genome and the maturity of the reference assembly play a role in the imputation efficiency, as indicated by the greater efficiency obtained in rice compared with alfalfa. While Beagle was preferable for rice, general data imputation methods performed significantly better in the absence of a reference genome. In particular, in alfalfa RFI and KNNI showed the lowest imputation errors in all classes, with KNNI to be preferred for computational efficiency. The alignment of markers to closely related species can help, but comes at the price of discarding markers that do not align, and with general methods still performing better.\\
The imputation accuracy tended to be relatively robust over increasing missing rates. However, KNNI showed a somewhat lower accuracy on high missingness scenarios, probably as a consequence of the ``curse of dimensionality''. In terms of computation requirements, all methods proved to be tractable over all problem complexities with a standard bioinformatics-lab computation infrastructure, except RFI (whose computational requirements increased exponentially with problem size and took up to 40 days with the complete rice dataset and maximum missing rate thresholds).

@@

\makeatletter{}\begin{acknowledgements}
The rice data used in this research paper were produced within the framework of the Italian national project ``RISINNOVA'' (grant n. 2010-2369), financially supported by the AGER Foundation.
The creation of the alfalfa data sets was funded by the projects ‘Genomic selection in alfalfa (GENALFA)’ funded by the Italian Ministry of Foreign Affairs and International Cooperation in the framework of the Italy-USA scientific cooperation program, the Italian share of the FP7-ArimNet project ‘Resilient, water- and energy-efficient forage and feed crops for Mediterranean agricultural systems (REFORMA)’ funded by the Italian Ministry of Agricultural and Forestry Policies.
\end{acknowledgements}
